---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

Hypothesis 2 Testing

HA: The species richness in Anopheles subgenera observed in African regions is decreasing
from 2007 to 2016.

H0: There is no change over time in the species richness within the Anopheles subgenera
observed in African regions.

Prediction: If we observe that mosquitoes within the Anopheles subgenus are expanding their
range latitudinally, we could attribute this effect to climate warming. The literature supports that climate warming would have a more significant effect on temperate regions (with higher latitude) and this can influence the distribution of Anopheline mosquitoes (Afrane et al., 2012; Ryan et al., 2020). As such, this change in latitude can exert selection pressure on the Anopheles subgenus, which could lead to a decrease in species richness as one or several subgenera may be more effective malaria vectors.

Data Analysis Plan: A mixed effect model will be used to test hypothesis 2. “Species Richness” will be the response
variable, and “time” (Year) will be included as a fixed effect. “Country” will be included as a
random effect, and we allow the slope to vary between different countries.
Model output should help us to determine whether the correlation between Species Richness
and Time is statistically significant.

```{r}
library(lme4)
library(lmerTest)
library(sjmisc)
library(MuMIn)
library(knitr)
```

```{r}
cases_dataset <- read.csv("C:/Users/sarah/Downloads/cases_dataset.csv")
```

#modelling species richness by country 
```{r}
mixed_model_sr_v_c <- lmer(Species_Richness ~ Year+(1|Country), data = cases_dataset, REML=T)
```

#only modelling the effect of country 
```{r}
mixed_model_c <- lmer(Species_Richness ~ 1+(1|Country), data = cases_dataset, REML=F)
summary(mixed_model_c)
```

#modelling the effect of country and year
```{r}
mixed_model_country <- lmer(Species_Richness ~ 1+(1|Country), data = cases_dataset, REML=F)

mixed_model_richness_year_country <- lmer(Species_Richness ~ Year+(1|Country), data = cases_dataset, REML=T)

mixed_model_int_country_year <- lm(Species_Richness ~ Year*Country, data = cases_dataset)
```

Creating visual
```{r}
names <- c("Year + (1|Country)", "(1|Country)", "Year * Country")

df <- c(4, 3, 81)

Rsqr <- c(r.squaredGLMM(mixed_model_richness_year_country)[2],
          r.squaredGLMM(mixed_model_country)[2],
          r.squaredGLMM(mixed_model_int_country_year)[2])

AICc <- c(1080.13, 1151.69, 931.33)

table <- data.frame(names, df, Rsqr, AICc)

kable_table <- kable(table, col.names = c("Candidate Models", "DF", "$R^2$", "$AIC_c$"), caption = "Ignore me")
kable_table
kable_table2 <- kable(table, col.names = c("Candidate Models", "DF", "$R^2$", "$AIC_c$"), caption = "Reporting Mixed Model Statistics")
kable_table2
```







































